---
title: "R Notebook"
output: html_notebook
---

Import Data

```{r}
mydata <- read.csv(file="D:/Northwestern/MSDS410.2/Week 8/RELIGION.csv",head=TRUE,sep=",")
religion <- data.frame(mydata)
str(religion)

```




1.   For the 2x2 table, determine the odds and the probabilities of texting while driving among males and females.  Then compute the odds ratio of texting while driving that compares males to females.  (5 points)

```{r}
male.yes <- 30
male.no <- 10
female.yes <- 34
female.no <- 6

#male odds of texting while driving
male.odds <- male.yes/male.no

#female odds of texting while driving
female.odds <- female.yes/female.no

#Male probability of texting while driving
male.prob <- male.yes/(male.yes+male.no)

#female probability of texting while driving
female.prob <- female.yes/(female.yes+female.no)

#odds ratio of texting while driving that compares males to females
odds.ratio <- (male.yes*female.no)/(male.no*female.yes)

```

a. Compute the overall odds and probability of attending a religious school, assuming this data is from a random sample.

```{r}
RS.table <- table(religion$RELSCHOL)
RS.table

prop.table(RS.table) # cell percentages
prop.table(RS.table, 1) # row percentages
prop.table(RS.table, 2) # column percentages

RS.odds <- 80/546
RS.odds

RS.prob <- 80/(546+80)
RS.prob

```

b. Cross-tabulate RELSCHOL with RACE (coded:  0=non-white, 1=white).  What are the probabilities that non-white students and white students attend religious schools?  What are the odds that white students and non-white students attend religious schools?  What is the odds ratio that compares white and non-white students?

```{r}

# 2-Way Cross Tabulation
library(gmodels)
CrossTable(religion$RELSCHOL, religion$RACE)
help(CrossTable)

xtabs(data=religion, ~RELSCHOL + RACE)
mytable<- table(religion$RELSCHOL, religion$RACE)
mytable
prop.table(mytable)

(26+54)/(26+54+76+470) #probability white and non-white
54/(54+470) #prob whites go to religious school
26/(26+76) #prob non-whites go to religious school

(26+54)/(76+470) #odds white and non-white
54/470 #odds whites go to religious school
26/(76) #odds non-whites go to religious school
(54*76)/(26*470) #odds ratio that compares whites and non-whites


```
Plot RELSCHOL (Y) by INCOME as a scatterplot.  The INCOME variable is actually an ordinal variable that is associated with income brackets.  This is an old dataset, so for example, INCOME=4 ??? $20,000-$29,999.   Is there a value of INCOME that seems to separate or discriminate between those attending religious schools and those that don???t?   Create a variable that dichotomizes INCOME based on this value you observed.  Call this new variable D_INCOME.  Cross-tabulate RELSCHOL with D_INCOME.  What are the probabilities that low income students and higher students attend religious schools?  What are the odds that lower income students and higher income students attend religious schools?  What is the odds ratio that compares lower and higher income students?  

```{r}
plot(y=religion$RELSCHOL, x=religion$INCOME)

mytable1 <-table(religion$RELSCHOL, religion$INCOME)
mytable1
prop.table(mytable1,1) # cell percentages
plot(prop.table(mytable1,2))

mytable2 <-table(religion$INCOME, religion$RELSCHOL)
mytable2
prop.table(mytable2,1) # cell percentages
plot(prop.table(mytable2,2))

religion$D_INCOME <- ifelse(religion$INCOME >3, 1, 0)
table(religion$RELSCHOL, religion$D_INCOME)
xtabs(data=religion, ~RELSCHOL + D_INCOME)

6/(6+131) #prob low income students attend religious schools
70/(70+383) #prob high income students attend religious schools
6/131 #odds low income students attend religious schools
70/383 #odds high income students attend religious schools




odds.ratio <- (6*383)/(131*70) #odds ratio that compares lower and higher income students
odds.ratio

library(lessR)
ANOVA(data=religion, INCOME ~ RELSCHOL)
ttest(data=religion, RELSCHOL ~ INCOME)


```

Plot RELSCHOL (Y) by ATTEND as a scatterplot.  The ATTEND variable is the number of times the survey respondent attends a service during a month.  Cross-tabulate RELSCHOL with ATTEND.  Are the proportion profiles the same for those attending religious school versus not, across the values of the ATTEND variable?  Is there a value of ATTEND that seems to separate or discriminate between those attending religious schools and those that don???t?   Save this value for later.

```{r}
plot(y=religion$RELSCHOL, x=religion$ATTEND)

mytable3 <-table(religion$RELSCHOL, religion$ATTEND)
mytable3
prop.table(mytable3,1) # cell percentages
plot(prop.table(mytable3,2))

mytable4 <-table(religion$ATTEND, religion$RELSCHOL)
mytable4
prop.table(mytable4,1) # cell percentages
prop.table(mytable4,2) # cell percentages
plot(prop.table(mytable4,2))

xtabs(data=religion, ~RELSCHOL + ATTEND)
mytable<- table(religion$RELSCHOL, religion$ATTEND)
mytable
prop.table(mytable)
Histogram(mytable)



```
3.   First, fit a logistic model to predict RELSCHOL (Y) using only the RACE (X) variable.  Call this Model 1.   Report the logistic regression model and interpret the parameter estimates for Model 1.  Report the AIC and BIC values for Model 1.    (3 points)

```{r}
Model1 <- Logit(RELSCHOL ~ RACE, family = binomial, data=religion)
Model1
anova(Model1)
summary(Model1)


Model1 <- glm(RELSCHOL ~ RACE, family = binomial, data=religion)
Model1
anova(Model1)
summary(Model1)
AIC(Model1)
BIC(Model1)
confint(Model1)
exp(coef(Model1))
exp(cbind(OR=coef(Model1),confint(Model1)))
religion$Model1.pred <- predict(Model1, newdata = religion, type="response")
religion$Model1.pred.round <- round(religion$model1.pred,0)
cm <- xtabs(~RELSCHOL + Model1.pred.round, data=religion)
cm
accuracy <- sum(diag(cm))/sum(cm)
accuracy

library(gmodels)
CrossTable(religion$model1.pred.round, religion$RELSCHOL,digits=1,prop.r=F, prop.t=F, prop.chisq=F, format="SPSS")

intercept <- -1.072637
b1 <- -1.091112

#prediction for whites
exp(intercept+b1*1)/(1+exp(intercept+b1*1))

#prediction for nonwhites
exp(intercept+b1*0)/(1+exp(intercept+b1*0))

#odds ratio
exp(-1.091112)

```
4.  Next, fit a logistic model to predict RELSCHOL (Y) using only the INCOME(X) variable.   Call this Model 2.   For Model 2, do the following:  (6 points)
a.	Report the logistic regression model and interpret the parameter estimates for Model 2.  Report the AIC and BIC values for Model 2.   How do these compare to Model 1?
b)  	Use the logit predictive equation for Model 2 to compute PI for each record.   Plot PI (Y) by INCOME(X).   At what value of X, does the value of PI exceed 0.50?   How does this value compare to your visual estimate from problem 2c)?   

```{r}
Model2 <- glm(RELSCHOL ~ INCOME, family = binomial, data=religion)
Model2
anova(Model2)
summary(Model2)
plot(Model2)
AIC(Model2)
BIC(Model2)

confint(Model2)
exp(coef(Model2))
exp(cbind(OR=coef(Model2),confint(Model2)))
religion$Model2.pred <- predict(Model2, newdata = religion, type="response")
religion$Model2.pred.round <- round(religion$Model2.pred,0)
cm <- xtabs(~RELSCHOL + Model2.pred.round, data=religion)
cm
accuracy <- sum(diag(cm))/sum(cm)
accuracy

library(gmodels)
CrossTable(religion$model2.pred.round, religion$RELSCHOL,digits=1,prop.r=F, prop.t=F, prop.chisq=F, format="SPSS")

intercept <- -2.82110
b1 <- 0.16228

#prediction for Income 1
exp(intercept+b1*1)/(1+exp(intercept+b1*1))

#prediction for Income 2
exp(intercept+b1*2)/(1+exp(intercept+b1*2))

#prediction for Income 3
exp(intercept+b1*3)/(1+exp(intercept+b1*3))

#prediction for Income 4
exp(intercept+b1*4)/(1+exp(intercept+b1*4))

#prediction for Income 5
exp(intercept+b1*5)/(1+exp(intercept+b1*5))

#prediction for Income 6
exp(intercept+b1*6)/(1+exp(intercept+b1*6))

#prediction for Income 7
exp(intercept+b1*7)/(1+exp(intercept+b1*7))

#prediction for Income 8
exp(intercept+b1*8)/(1+exp(intercept+b1*8))

#prediction for Income 9
exp(intercept+b1*9)/(1+exp(intercept+b1*9))

#prediction for Income 10
exp(intercept+b1*10)/(1+exp(intercept+b1*10))

#prediction for Income 11
exp(intercept+b1*11)/(1+exp(intercept+b1*11))

#prediction for Income 12
exp(intercept+b1*12)/(1+exp(intercept+b1*12))

#odds ratio
exp(0.16228)

library(ggplot2)

ggplot(religion, aes(x=INCOME,y= Model2.pred))+
  geom_point(col="dark green", fill="green")+
  ggtitle("Scatter Plot of RELSCHOL prediction given INCOME")+
  xlab("INCOME")+
  ylab("Model2.pred")+
  xlim(0, 12)+
  ylim(0, 1)+
  scale_x_discrete(limits=c(1,2,3,4,5,6,7,8,9,10,11,12))
  

```

5.  Next, fit a logistic model to predict RELSCHOL (Y) using only the ATTEND(X) variable.   Call this Model 3.   For Model 3, do the following:  (6 points)

```{r}
Model3 <- glm(RELSCHOL ~ ATTEND, family = binomial, data=religion)
Model3
anova(Model3)
summary(Model3)
plot(Model3)
AIC(Model3)
BIC(Model3)

confint(Model3)
exp(coef(Model3))
exp(cbind(OR=coef(Model3),confint(Model3)))
religion$Model3.pred <- predict(Model3, newdata = religion, type="response")
religion$Model3.pred.round <- round(religion$Model3.pred,0)
cm <- xtabs(~RELSCHOL + Model3.pred.round, data=religion)
cm
accuracy <- sum(diag(cm))/sum(cm)
accuracy

library(gmodels)
CrossTable(religion$model3.pred.round, religion$RELSCHOL,digits=1,prop.r=F, prop.t=F, prop.chisq=F, format="SPSS")

intercept <- -2.9727
b1 <- 0.2269 

#prediction for ATTEND1
exp(intercept+b1*1)/(1+exp(intercept+b1*1))

#prediction for ATTEND2
exp(intercept+b1*2)/(1+exp(intercept+b1*2))

#prediction for ATTEND3
exp(intercept+b1*3)/(1+exp(intercept+b1*3))

#prediction for ATTEND4
exp(intercept+b1*4)/(1+exp(intercept+b1*4))

#prediction for ATTEND5
exp(intercept+b1*5)/(1+exp(intercept+b1*5))

#prediction for ATTEND6
exp(intercept+b1*6)/(1+exp(intercept+b1*6))


#odds ratio
exp(0.16228)

library(ggplot2)

ggplot(religion, aes(x=ATTEND,y= Model3.pred))+
  geom_point(col="dark green", fill="green")+
  ggtitle("Scatter Plot of RELSCHOL prediction given ATTEND")+
  xlab("INCOME")+
  ylab("Model3.pred")+
  xlim(0, 12)+
  ylim(0, 1)+
  scale_x_discrete(limits=c(1,2,3,4,5,6,7,8,9,10,11,12))
  

```

6.  Finally, fit a logistic model to predict RELSCHOL (Y) using RACE, INCOME and ATTEND as explanatory (X) variables.  Please consider INCOME and ATTEND to be continuous variables.  Call this Model 4.   For Model 4, do the following:  (9 points)
```{r}
Model4 <- glm(RELSCHOL ~ RACE + INCOME + ATTEND, family = binomial, data=religion)
Model4
anova(Model4)
summary(Model4)
plot(Model4)
AIC(Model4)
BIC(Model4)

confint(Model4)
exp(coef(Model4))
exp(cbind(OR=coef(Model4),confint(Model4)))
religion$Model4.pred <- predict(Model4, newdata = religion, type="response")
religion$Model4.pred.round <- round(religion$Model4.pred,0)
cm <- xtabs(~RELSCHOL + Model4.pred.round, data=religion)
cm
accuracy <- sum(diag(cm))/sum(cm)
accuracy

intercept <- -3.58314
b1 <- -1.28931 
b2 <- 0.20068
b3 <- 0.33164

#prediction for Whites with income level 4 who attend 5 religious services per month
exp(intercept+b1*1+b2*4+b3*5)/(1+exp(intercept+b1*1+b2*4+b3*5))

4/45

#prediction for non-Whites with income level 4 who attend 5 religious services per month
exp(intercept+b1*0+b2*4+b3*5)/(1+exp(intercept+b1*0+b2*4+b3*5))

#OR white vs non-white
exp(-1.28931)
```

